---
title: "balanced_nonoverlapping_flash_normalf"
author: "Annie Xie"
date: "2025-02-04"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this analysis, we are exploring EBMF with a normal prior on F in the balanced, nonoverlapping setting.

```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(pheatmap)
library(flashier)
```

```{r}
source('code/visualization_functions.R')
```

```{r}
compute_L2_fit <- function(est, dat, with_diag = FALSE){
  if (with_diag == FALSE){
    score <- sum((dat - est)^2) - sum((diag(dat) - diag(est))^2)
  }
  else{
    score <- sum((dat - est)^2)
  }
  return(score)
}
```

# Flash with Normal Prior on F (with backfitting)

```{r}
group_nonoverlap_1 <- readRDS("data/group_nonoverlap_1.rds")
group_nonoverlap_1_flash_normalf_1 <- readRDS("data/group_nonoverlap_1_flash_normalf_1.rds")
```

## Visualization of Estimate
This is a heatmap of the true $L$ that we are hoping to recover:
```{r}
plot_heatmap(group_nonoverlap_1$true_L)
```

This is a heatmap of $\hat{L}$, the estimate for $L$ from Flashier with normal prior on $F$:
```{r}
plot_heatmap(group_nonoverlap_1_flash_normalf_1$est_L, brks = seq(0, max(group_nonoverlap_1_flash_normalf_1$est_L), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(group_nonoverlap_1_flash_normalf_1$est_L, pops_vec)
```

## Visualizations related to fit

This is the true Gram matrix, $\frac{1}{p}XX'$:
```{r}
plot_heatmap(group_nonoverlap_1$data$YYt, colors = c('blue','gray96', 'red'), brks = seq(-max(abs(group_nonoverlap_1$data$YYt)), max(abs(group_nonoverlap_1$data$YYt)), length.out = 50))
```

This is a heatmap of the estimate of the Gram matrix, $\hat{L}\hat{L}'$:
```{r}
plot_heatmap(group_nonoverlap_1_flash_normalf_1$est_LLt, brks = seq(0, max(group_nonoverlap_1_flash_normalf_1$est_LLt), length.out = 50))
```

This is a scatter plot of the fitted values vs observed values:
```{r}
ggplot(data = NULL, aes(x = c(group_nonoverlap_1$data$YYt), y = c(group_nonoverlap_1_flash_normalf_1$est_LLt))) + geom_point() + ylim(-1, 9) + xlim(-1,9) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

Let's focus on the off-diagonal elements:
```{r}
diag_idx <- seq(1, prod(dim(group_nonoverlap_1$data$YYt)), length.out = ncol(group_nonoverlap_1$data$YYt))
off_diag_idx <- setdiff(c(1:prod(dim(group_nonoverlap_1$data$YYt))), diag_idx) 
```

```{r}
ggplot(data = NULL, aes(x = c(group_nonoverlap_1$data$YYt)[off_diag_idx], y = c(group_nonoverlap_1_flash_normalf_1$est_LLt)[off_diag_idx])) + geom_point() + ylim(-1, 9) + xlim(-1,9) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

As mentioned in other explorations, ideally this scatter plot would have two clusters -- one cluster for the near-zero entries (corresponding to the sparse entries) and another cluster corresponding to the group effects. In this plot, we see that the fitted values corresponding to near-zero observed values range from 1-3., representing an elevated baseline. We have one cluster which falls on the y=x line; this seems to correspond to the group effect for group 3 (the group that was in the intercept factor). Lastly, we see a cluster with fitted values around 7.5-8; this corresponds to the group effect factors for the other groups. 

This is the elbo of the flash fit:
```{r}
group_nonoverlap_1_flash_normalf_1$fit_obj$elbo
```

## Observations
Similar to EBMFcov-greedy in this setting, Flashier with a normal prior on $F$ seems to add something like an intercept factor. In the ``intercept" factor, the group effect for one group is more prominent than the others; samples from this group have a higher loading value than others. This could be due to the backfitting step; it's possible that Flashier with a normal prior on $F$ without backfitting would lead to an estimate similar to that of EBMFcov-greedy (this is something we check in a later section). Another thought is whether backfitting would make the intercept factor more sparse, turning it into a group effect factor. 

This estimate of $L$ doesn't lead to the best estimate for the Gram matrix. Due to the intercept-like factor, the estimate for the Gram matrix is not sparse.

# Try additional backfitting

```{r}
flash_normalf_fit <- group_nonoverlap_1_flash_normalf_1$fit_obj
flash_normalf_fit %>%
  flash_backfit(tol = 10^(-16), verbose = 3)
```

## Visualizations of Estimate
This is a heatmap of $\hat{L}_{backfit}$, the estimate for $L$:
```{r}
plot_heatmap(flash_normalf_fit$L_pm, brks = seq(0, max(flash_normalf_fit$L_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}_{backfit}$:
```{r}
plot_loadings(group_nonoverlap_1_flash_normalf_1$est_L, pops_vec)
```

This is the elbo:
```{r}
flash_normalf_fit$elbo
```

## Observations
One thing I did was try increasing th number of maximum backfitting iterations. Increasing the number of maximum backfitting iterations does not change the estimate. (I checked the number of iterations afterwards and it only took 28 backfitting iterations to stop, so that makes sense). Another thing I tried was decreasing the tolerance level. This also does not change the estimate. Therefore, for this setting, we can eliminate the possibility that the backfitting is converging slowly. 

# For comparison, Flash with normal prior on F with no backfitting
```{r}
flash_normalf_nobackfit.wrapper <- function(input, args){
  Kmax <- input$K
  flash_fit <- flash_init(data = input$Y, var_type = 0) %>%
    flash_greedy(ebnm_fn = c(args$ebnm_fn, ebnm::ebnm_normal), Kmax = Kmax)
  flash_ldf <- ldf(flash_fit, type = "f") # I think the default option is type = f
  flash_scaled_L <- (1/sqrt(ncol(input$Y)))*flash_ldf$L %*% diag(flash_ldf$D)
  return(list(flash_fit = flash_fit, est_LLt = tcrossprod(flash_scaled_L)))
}
flash_normalf_nobackfit_data <- flash_normalf_nobackfit.wrapper(group_nonoverlap_1$data, args = list(ebnm_fn = ebnm::ebnm_generalized_binary))
```

## Visualization of Estimate
This is a heatmap of $\hat{L}_{greedy}$, the estimate for $L$ from Flashier with normal prior on $F$ with no backfit:
```{r}
plot_heatmap(flash_normalf_nobackfit_data$flash_fit$L_pm, brks = seq(0, max(flash_normalf_nobackfit_data$flash_fit$L_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}_{greedy}$:
```{r}
plot_loadings(flash_normalf_nobackfit_data$flash_fit$L_pm, pops_vec)
```

This is the elbo:
```{r}
flash_normalf_nobackfit_data$flash_fit$elbo
```

## Observations
The estimate from Flashier with normal prior on F without backfitting looks like that from EBMFcov (without backfitting). The first factor from Flashier can often be interpreted as an intercept factor, so this behavior is not too surprising to me. This is also similar to behavior noted in the coordinate descent symnmf paper by Vandaele et al. (they note how if you use zero as an initial estimate for L, the first factor is often dense and the others are more sparse).

# Try initializing with true values
```{r}
flash_normalf_fit_true_init <- flash_init(group_nonoverlap_1$data$Y) %>%
  flash_factors_init(init = list(group_nonoverlap_1$data$LL, group_nonoverlap_1$data$FF), ebnm_fn = c(ebnm::ebnm_generalized_binary, ebnm::ebnm_normal)) %>%
  flash_backfit()
```

## Visualizations of the Estimate
This is a heatmap of $\hat{L}_{true-init}$:
```{r}
plot_heatmap(flash_normalf_fit_true_init$L_pm, brks = seq(0, max(flash_normalf_fit_true_init$L_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}_{true-init}$:
```{r}
plot_loadings(flash_normalf_fit_true_init$L_pm, pops_vec)
```

This is the corresponding elbo:
```{r}
flash_normalf_fit_true_init$elbo
```

## Observations
We see that when we initialize at the true values, the estimate of $L$ looks much closer to the true value. The elbo is also higher, meaning this estimate is a better estimate (with respect to the objective function) than the one found with the greedy initialization. I'm curious how the joint-update algorithm compares to this.

# Analyzing fit of the data matrix

## Visualizations related to fit
These visualizations are related to the fit of the data matrix, $X$.

This is a scatter plot of the fitted values for $X$ vs the entries of $X$:
```{r}
X_fitted_vals <- fitted(group_nonoverlap_1_flash_normalf_1$fit_obj)
```

```{r}
ggplot(data = NULL, aes(x = c(group_nonoverlap_1$data$Y), y = c(X_fitted_vals))) + geom_point() + xlim(-11, 11) + ylim(-11, 11) + xlab('entries of X') + ylab('fitted values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

## Visualizations of the first component
These are visualizations of one of the components, $l_k f_k'$, from the decomposition of $X$.
```{r}
est_component1 <- (group_nonoverlap_1_flash_normalf_1$fit_obj$L_pm[,1] %*% t(group_nonoverlap_1_flash_normalf_1$fit_obj$F_pm[,1]))
```

```{r}
true_component1 <- group_nonoverlap_1$data$LL[,3] %*% t(group_nonoverlap_1$data$FF[,3])
```

This is a heatmap of the first estimated component:
```{r}
plot_heatmap(est_component1, colors_range = c('blue', 'gray96','red'), brks = seq(-max(abs(est_component1)), max(abs(est_component1)), length.out = 50))
```

This is a scatter plot of one column from the first estimated component:
```{r}
ggplot(data = NULL, aes(x = c(1:160))) + geom_point(aes(y = c(true_component1[,1])), color = 'red', alpha = 0.5) + geom_point(aes(y = c(est_component1[,1])), color = 'blue', alpha = 0.5) + xlab('index') + ylab('component entry')
```

This is a heatmap of the residualized $X$ matrix (this would be the data matrix when re-fitting the first component during backfitting):
```{r}
X_resid_est1 <- group_nonoverlap_1$data$Y - (group_nonoverlap_1_flash_normalf_1$fit_obj$L_pm[,2:4] %*% t(group_nonoverlap_1_flash_normalf_1$fit_obj$F_pm[,2:4]))
```

```{r}
plot_heatmap(X_resid_est1, colors_range = c('blue', 'gray96','red'), brks = seq(-max(abs(X_resid_est1)), max(abs(X_resid_est1)), length.out = 50))
```

This is a heatmap of $X$ residualized using the true values for the components:
```{r}
X_resid_true1 <- group_nonoverlap_1$data$Y - (group_nonoverlap_1$data$LL[,c(1,2,4)] %*% t(group_nonoverlap_1$data$FF[,c(1,2,4)]))
```

```{r}
plot_heatmap(X_resid_true1, colors_range = c('blue', 'gray96','red'), brks = seq(-max(abs(X_resid_true1)), max(abs(X_resid_true1)), length.out = 50))
```

## Visualizations related to F

This is a heatmap of $\hat{F}'\hat{F}$:
```{r}
est_FtF <- crossprod(group_nonoverlap_1_flash_normalf_1$fit_obj$F_pm)
```

```{r}
plot_heatmap(est_FtF, colors_range = c('blue', 'gray96','red'), brks = seq(-max(abs(est_FtF)), max(abs(est_FtF)), length.out = 50))
```

This is a heatmap of $\hat{L}(\hat{F}'\hat{F})$:
```{r}
LFtF_est <- group_nonoverlap_1_flash_normalf_1$est_L %*% est_FtF
```

```{r}
plot_heatmap(LFtF_est, colors_range = c('blue', 'gray96','red'), brks = seq(-max(abs(LFtF_est)), max(abs(LFtF_est)), length.out = 50))
```

This is a scatter plot of one of the columns of $\hat{L}(\hat{F}'\hat{F})$:
```{r}
plot(LFtF_est[,1])
```

## Observations
Analyzing the fit of the observations to the data matrix, the fitted values match the observed values relatively well. When considering $LF'FL'$ as an estimate for $XX'$ (as opposed to $LL'$), it seems like the negative entries in $F'F$ are cancelling out the baseline non-zero entries in the first factor of $L$.

# Try more factors
Another question of interest is whether adding a fifth factor would add the final group effect. Furthermore, if we do backfit, then it would make sense for the first factor to zero out (since it is not necessary to explain the fit anymore). We test this out in this section.

```{r}
flash_normalf_K5_fit <- flash_init(group_nonoverlap_1$data$Y) %>%
  flash_greedy(Kmax = 5, ebnm_fn = c(ebnm::ebnm_generalized_binary, ebnm_normal)) %>%
  flash_backfit() %>%
  flash_nullcheck()
```

## Visualization of Estimate

This is a heatmap of $\hat{L}$, the estimate for $L$ from Flashier with normal prior on $F$:
```{r}
plot_heatmap(flash_normalf_K5_fit$L_pm, brks = seq(0, max(flash_normalf_K5_fit$L_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}_{K5}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(flash_normalf_K5_fit$L_pm, pops_vec)
```

This are the entries of the first column of $\hat{L}_{K5}$:
```{r}
flash_normalf_K5_fit$L_pm[1,1]
```

This is the elbo of the flash fit:
```{r}
flash_normalf_K5_fit$elbo
```

## Visualizations related to fit

This is a heatmap of the estimate of the Gram matrix, $\hat{L}_{K5}\hat{L}_{K5}'$:

```{r}
scaled_flash_normalf_K5_fit <- ldf(flash_normalf_K5_fit, type = 'f')
scaled_flash_normalf_K5_estLLt <- (1/ncol(group_nonoverlap_1$data$Y)) * (scaled_flash_normalf_K5_fit$L %*% diag(scaled_flash_normalf_K5_fit$D^2) %*% t(scaled_flash_normalf_K5_fit$L))
```

```{r}
plot_heatmap(scaled_flash_normalf_K5_estLLt, brks = seq(0, max(scaled_flash_normalf_K5_estLLt), length.out = 50))
```

This is a scatter plot of the fitted values vs observed values:
```{r}
ggplot(data = NULL, aes(x = c(group_nonoverlap_1$data$YYt), y = c(scaled_flash_normalf_K5_estLLt))) + geom_point() + ylim(-1, 9) + xlim(-1,9) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

Let's focus on the off-diagonal elements:
```{r}
diag_idx <- seq(1, prod(dim(group_nonoverlap_1$data$YYt)), length.out = ncol(group_nonoverlap_1$data$YYt))
off_diag_idx <- setdiff(c(1:prod(dim(group_nonoverlap_1$data$YYt))), diag_idx) 
```

```{r}
ggplot(data = NULL, aes(x = c(group_nonoverlap_1$data$YYt)[off_diag_idx], y = c(scaled_flash_normalf_K5_estLLt)[off_diag_idx])) + geom_point() + ylim(-1, 9) + xlim(-1,9) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

This is a heatmap of $\hat{F}_{K5}'\hat{F}_{K5}$:
```{r}
scaled_est_FtF_K5 <- crossprod(scaled_flash_normalf_K5_fit$F)
```

```{r}
plot_heatmap(scaled_est_FtF_K5, colors_range = c('blue', 'gray96','red'), brks = seq(-max(abs(scaled_est_FtF_K5)), max(abs(scaled_est_FtF_K5)), length.out = 50))
```

## Observations
Setting `Kmax = 5` does add a factor that captures the last group effect. This makes sense since the rows of the data matrix are essentially the group effect factors minus some mean value. So there is a subset of rows that still contain the last group effect. The backfitting does decrease the loading value of the first factor (the intercept factor) to some very small number (2.6e-06). This also makes sense because once we have the four group effect factors, the intercept factor is not needed (I did not use an intercept term when simulating this data). 

# Work in Progress - Testing how many factors it wants to add
In this section, my goal is to determine how many factors flashier wants to add.

## Just Greedy
```{r}
flash_normalf_largeK_fit <- flash_init(group_nonoverlap_1$data$Y) %>%
  flash_greedy(Kmax = 10, ebnm_fn = c(ebnm::ebnm_generalized_binary, ebnm_normal)) %>%
  flash_nullcheck()
```

This is a heatmap of $\hat{L}_{large-K}$:
```{r}
plot_heatmap(flash_normalf_largeK_fit$L_pm, brks = seq(0, max(flash_normalf_largeK_fit$L_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}_{large-K}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(flash_normalf_largeK_fit$L_pm, pops_vec)
```

This is the L2 norm of the columns of the factor matrix, $\hat{F}_{large-K}$:
```{r}
apply(flash_normalf_largeK_fit$F_pm, 2, function(x){sqrt(sum(x^2))})
```

```{r}
scaled_flash_normalf_largeK_fit <- ldf(flash_normalf_largeK_fit)
scaled_flash_normalf_largeK_L <- scaled_flash_normalf_largeK_fit$L %*% diag(scaled_flash_normalf_largeK_fit$D)
```

This is the elbo:
```{r}
flash_normalf_largeK_fit$elbo
```

This is the L2 norm of the difference between the observed Gram matrix and estimated Gram matrix (not including diagonal entries):
```{r}
compute_L2_fit((1/ncol(group_nonoverlap_1$data$Y))*tcrossprod(scaled_flash_normalf_largeK_L), group_nonoverlap_1$data$YYt)
```

This is the L2 norm of the difference between the observed Gram matrix and estimated Gram matrix (including diagonal entries):
```{r}
compute_L2_fit((1/ncol(group_nonoverlap_1$data$Y))*tcrossprod(scaled_flash_normalf_largeK_L), group_nonoverlap_1$data$YYt, with_diag = TRUE)
```

This is the estimate of the residual standard deviation:
```{r}
flash_normalf_largeK_fit$residuals_sd
```

This is the KL term:
```{r}
sum(unlist(flash_normalf_largeK_fit$flash_fit$KL)) # it seems like the KL changes the most
```

This is the KL for the individual loadings and factors:
```{r}
flash_normalf_largeK_fit$flash_fit$KL # it seems like the KL changes the most
```

This is the estimate for the prior for the sixth loadings column:
```{r}
flash_normalf_largeK_fit$L_ghat[[6]]
```

This is a heatmap of the residualized data matrix:
```{r}
X_resid_est <- group_nonoverlap_1$data$Y - fitted(flash_normalf_largeK_fit)
```

```{r}
plot_heatmap(X_resid_est, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(X_resid_est)), max(abs(X_resid_est)), length.out = 50))
```

This is a scatterplot of one of the columns of the residualized data matrix:
```{r}
plot(X_resid_est[,4])
```

## Add Backfitting
```{r}
flash_normalf_largeK_backfit_fit <- flash_normalf_largeK_fit %>% flash_backfit() %>% flash_nullcheck()
```

This is the number of factors kept in the backfit:
```{r}
flash_normalf_largeK_backfit_fit$n_factors
```

This is a heatmap of $\hat{L}$, the estimate for $L$ from Flashier with normal prior on $F$:
```{r}
plot_heatmap(flash_normalf_largeK_backfit_fit$L_pm, brks = seq(0, max(flash_normalf_largeK_backfit_fit$L_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}_{K}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(flash_normalf_largeK_backfit_fit$L_pm, pops_vec)
```

This is the elbo:
```{r}
flash_normalf_largeK_backfit_fit$elbo
```

```{r}
scaled_flash_normalf_largeK_backfit_fit <- ldf(flash_normalf_largeK_backfit_fit)
scaled_flash_normalf_largeK_backfit_L <- scaled_flash_normalf_largeK_backfit_fit$L %*% diag(scaled_flash_normalf_largeK_backfit_fit$D)
```

This is the L2 norm of the difference between the observed Gram matrix and estimated Gram matrix (not including diagonal entries):
```{r}
compute_L2_fit((1/ncol(group_nonoverlap_1$data$Y))*tcrossprod(scaled_flash_normalf_largeK_backfit_L), group_nonoverlap_1$data$YYt)
```

This is the L2 norm of the difference between the observed Gram matrix and estimated Gram matrix (including diagonal entries):
```{r}
compute_L2_fit((1/ncol(group_nonoverlap_1$data$Y))*tcrossprod(scaled_flash_normalf_largeK_backfit_L), group_nonoverlap_1$data$YYt, with_diag = TRUE)
```

This is the estimate of the residual standard deviation:
```{r}
flash_normalf_largeK_backfit_fit$residuals_sd
```

This is the KL term:
```{r}
sum(unlist(flash_normalf_largeK_backfit_fit$flash_fit$KL)) # it seems like the KL changes the most
```

This is the KL of the inidivudal loadings and factors:
```{r}
flash_normalf_largeK_backfit_fit$flash_fit$KL # it seems like the KL changes the most
```

This is the estimate of the prior for the sixth loadings vector:
```{r}
flash_normalf_largeK_backfit_fit$L_ghat[[6]]
```

This is a heatmap of the residualized data matrix:
```{r}
X_backfit_resid_est <- group_nonoverlap_1$data$Y - fitted(flash_normalf_largeK_backfit_fit)
```

```{r}
plot_heatmap(X_backfit_resid_est, colors_range = c('blue','gray96','red'), brks = seq(-max(abs(X_backfit_resid_est)), max(abs(X_backfit_resid_est)), length.out = 50))
```

## Observations
I tested setting `Kmax` to 100 and flashier did add 100 factors in the greedy procedure. Taking a closer look at the factors, we have an intercept factor, four group effect factors, and then a bunch of factors with a very small entry value. I tried backfitting on top of this fit. The entries of the intercept factor decrease to a very small value. The backfitting also found two factors to be numerically zero. However, it kept the other factors with very small entry values. I checked the corresponding factors in the $F$ matrix, and these columns also have very small entries (but not zero). 

I wonder if changing the tolerance settings would cause these factors to become zero vectors. I also find this behavior puzzling since it seems to keep adding a new intercept factor where all the entries are a very small value. If, hypothetically, it did need an intercept factor, why wouldn't it just add everything together in one intercept factor?

Based off my experimentation, it seems like the KL term is primarily what is changing when a new factor is being added. I find that the KL term associated with the new loadings factor is usually 55.45177 and the KL term associated with the new factor is usually a really small number. As a result, the new factor is able to decrease the magnitude of the negative KL term and increase the elbo. I'm still trying to figure out a narrative for why this is happening.

<!--
# Notes
This is a record of other things I've tried. 
1. Fitting flash with a fixed standard normal prior on F. I did not expect this to make a difference, and it, in fact, did not -- the estimate still recovered the same type of structure.

# Test: fixed prior
```{r, eval = FALSE, include = FALSE}
ebnm_std_normal_fixed <- function(x, s, g_init, fix_g, output) {
  ebnm_res <- ebnm::ebnm_normal(
    x = x,
    s = s,
    g_init = ashr::normalmix(pi = 1, mean = 0, sd = 1),
    fix_g = TRUE,
    output = output,
    control = list(iterlim = 10)
  )
  return(ebnm_res)
}
```

```{r, eval = FALSE, include = FALSE}
flash_fixed_normalf.wrapper <- function(input, args){
  Kmax <- input$K
  flash_fit <- flash_init(data = input$Y, var_type = 0) %>%
    flash_greedy(ebnm_fn = c(args$ebnm_fn, ebnm_std_normal_fixed), Kmax = Kmax) %>%
    flash_backfit()
  flash_ldf <- ldf(flash_fit, type = "f") # I think the default option is type = f
  flash_scaled_L <- (1/sqrt(ncol(input$Y)))*flash_ldf$L %*% diag(flash_ldf$D)
  return(list(flash_fit = flash_fit, est_LLt = tcrossprod(flash_scaled_L)))
}
flash_fixed_normalf_data <- flash_fixed_normalf.wrapper(group_nonoverlap_1$data, args = list(ebnm_fn = ebnm::ebnm_generalized_binary))
```
-->