---
title: "balanced_nonoverlapping_ebmfcov_diag"
author: "Annie Xie"
date: "2025-02-03"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this analysis, we are interested in further exploring the EBMFcov method in the balanced non-overlapping setting.

```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(pheatmap)
library(flashier)
```

```{r}
source('code/visualization_functions.R')
```

# EBMFcov (without backfitting)
## Visualizations of the estimate

```{r}
group_nonoverlap_1 <- readRDS("data/group_nonoverlap_1.rds")
group_nonoverlap_1_ebmfcov_diag_1 <- readRDS("data/group_nonoverlap_1_ebmfcov_diag_1.rds")
```

This is a heatmap of the true $L$ that we hope to recover:
```{r}
plot_heatmap(group_nonoverlap_1$true_L)
```

This is a heatmap of $\hat{L}$, the estimate for $L$ from EBMFcov:
```{r}
plot_heatmap(group_nonoverlap_1_ebmfcov_diag_1$est_L, brks = seq(0, max(group_nonoverlap_1_ebmfcov_diag_1$est_L), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(group_nonoverlap_1_ebmfcov_diag_1$est_L, pops_vec)
```

## Visualizations related to fit

This is the true Gram matrix, $\frac{1}{p}XX'$:
```{r}
plot_heatmap(group_nonoverlap_1$data$YYt, brks = seq(0, max(group_nonoverlap_1$data$YYt), length.out = 50))
```

This is a heatmap of the estimate of the Gram matrix, $\hat{L}\hat{L}'$:
```{r}
plot_heatmap(group_nonoverlap_1_ebmfcov_diag_1$est_LLt, brks = seq(0, max(group_nonoverlap_1_ebmfcov_diag_1$est_LLt), length.out = 50))
```

This is a scatter plot of the fitted values vs observed values:
```{r}
ggplot(data = NULL, aes(x = c(group_nonoverlap_1$data$YYt), y = c(group_nonoverlap_1_ebmfcov_diag_1$est_LLt))) + geom_point() + ylim(0, 6) + xlim(0,6) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

The model that we fit is $XX' \approx LL' + D$, so let's focus on the off-diagonal elements for now.
```{r}
diag_idx <- seq(1, prod(dim(group_nonoverlap_1$data$YYt)), length.out = ncol(group_nonoverlap_1$data$YYt))
off_diag_idx <- setdiff(c(1:prod(dim(group_nonoverlap_1$data$YYt))), diag_idx) 
```

```{r}
ggplot(data = NULL, aes(x = c(group_nonoverlap_1$data$YYt)[off_diag_idx], y = c(group_nonoverlap_1_ebmfcov_diag_1$est_LLt)[off_diag_idx])) + geom_point() + ylim(0, 6) + xlim(0,6) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

## Observations
Based off the visualizations, it seems like this method wants to add an intercept term. This is not totally surprising since the loadings and factors are added in a greedy way, and for this scenario, an intercept factor provides an equally good rank one fit as a factor identifying one group. However, looking at the heatmap of $LL'$, this looks like a bad estimate for the Gram matrix. (Side note: it's possible that if we add another factor, this fifth factor would pick up the last group.)

We wonder if backfitting would fix this issue because if you took out the effect of the three groups, then the intercept factor would be a bad fit to the residualized data matrix. We will test this in the next section.

# Adding backfitting
Now, we add a backfitting step whenever we fit the flash fits.
```{r}
cov_fit_backfit <- function(covmat, ebnm_fn = ebnm::ebnm_point_laplace, Kmax = 1000, verbose.lvl = 0, backfit_iter = 500) {
  fl <- flash_init(covmat, var_type = 0) %>%
    flash_set_verbose(verbose.lvl) %>%
    flash_greedy(ebnm_fn = ebnm_fn, Kmax = Kmax) %>%
    flash_backfit(maxiter = backfit_iter)
  s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
  s2_diff <- Inf
  while(s2 > 0 && abs(s2_diff - 1) > 1e-4) {
    covmat_minuss2 <- covmat - diag(rep(s2, ncol(covmat)))
    fl <- flash_init(covmat_minuss2, var_type = 0) %>%
      flash_set_verbose(verbose.lvl) %>%
      flash_greedy(ebnm_fn = ebnm_fn, Kmax = Kmax) %>%
      flash_backfit(maxiter = backfit_iter)
    old_s2 <- s2
    s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
    s2_diff <- s2 / old_s2
  }
  
  return(list(fl=fl, s2 = s2))
}
```

```{r}
cov_fit_backfit_fit <- cov_fit_backfit(group_nonoverlap_1$data$YYt, ebnm_fn = ebnm::ebnm_generalized_binary, 
                                       Kmax = group_nonoverlap_1$data$K, backfit_iter = 500)
```

## Visualization of Estimate

This is a heatmap of $\hat{L}_{backfit}$, the estimate for $L$ from EBMFcov with backfitting:
```{r}
plot_heatmap(cov_fit_backfit_fit$fl$L_pm, brks = seq(0, max(cov_fit_backfit_fit$fl$L_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}_{backfit}$:
```{r}
plot_loadings(cov_fit_backfit_fit$fl$L_pm, pops_vec)
```

This is the elbo of the flash fit:
```{r}
cov_fit_backfit_fit$fl$elbo
```

## Visualizations related to fit

This is a heatmap of $\hat{L}_{backfit} \hat{L}_{backfit}'$:
```{r}
cov_fit_backfit_fit_estLLt <- tcrossprod(cov_fit_backfit_fit$fl$L_pm)
```

```{r}
plot_heatmap(cov_fit_backfit_fit_estLLt, brks = seq(0, max(cov_fit_backfit_fit_estLLt), length.out = 50))
```

This is a scatter plot of the fitted values vs observed values:
```{r}
ggplot(data = NULL, aes(x = c(group_nonoverlap_1$data$YYt), y = c(cov_fit_backfit_fit_estLLt))) + geom_point() + ylim(0, 6) + xlim(0,6) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

This is a scatter plot of the off-diagonal entries of the fitted values vs those of the observed values:
```{r}
ggplot(data = NULL, aes(x = c(group_nonoverlap_1$data$YYt)[off_diag_idx], y = c(cov_fit_backfit_fit_estLLt)[off_diag_idx])) + geom_point() + ylim(0, 6) + xlim(0,6) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

In this scatter plot, we should ideally see two groups -- one corresponding to the near-zero values of the Gram matrix, and one corresponding to the non-zero values for samples within the same group. We see one group where the observed values are around 4-5 and the fitted values are around 1 -- this probably corresponds to the group effect that was recovered only after backfitting. This group effect has a lower loadings value than the other group effects.

## Observations
We see that with the backfit, the estimate looks closer to what we want. This suggests that this is a computational issue (i.e. the objective function is not actively favoring the first estimate) that can be mitigated with backfitting. However, the recovered group effect is not as strong as the other group effects. I tried backfitting with larger numbers of iterations to see if the estimate improves (i.e. testing does the backfit have slow convergence?), but it doesn't seem to improve the estimate.