---
title: "balanced_nonoverlapping_ebmfcov_diag"
author: "Annie Xie"
date: "2025-02-03"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this analysis, we are interested in further exploring the EBMFcov method in the balanced non-overlapping setting.

```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(pheatmap)
library(flashier)
```

```{r}
source('code/visualization_functions.R')
```

# EBMFcov (without backfitting)
## Visualizations of the estimate

```{r}
group_nonoverlap_1 <- readRDS("data/group_nonoverlap_1.rds")
group_nonoverlap_1_ebmfcov_diag_1 <- readRDS("data/group_nonoverlap_1_ebmfcov_diag_1.rds")
```

This is a heatmap of the true $L$ that we hope to recover:
```{r}
plot_heatmap(group_nonoverlap_1$true_L)
```

This is a heatmap of $\hat{L}$, the estimate for $L$ from EBMFcov:
```{r}
plot_heatmap(group_nonoverlap_1_ebmfcov_diag_1$est_L, brks = seq(0, max(group_nonoverlap_1_ebmfcov_diag_1$est_L), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(group_nonoverlap_1_ebmfcov_diag_1$est_L, pops_vec)
```

This is a heatmap of $\hat{F}$, the estimate for $F$ from EBMFcov (note, I did not scale this, so the scaling is arbitrary):
```{r}
plot_heatmap(group_nonoverlap_1_ebmfcov_diag_1$fit_obj$fl$F_pm, brks = seq(0, max(group_nonoverlap_1_ebmfcov_diag_1$fit_obj$fl$F_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(group_nonoverlap_1_ebmfcov_diag_1$fit_obj$fl$F_pm, pops_vec)
```

```{r}
group_nonoverlap_1_ebmfcov_diag_1$fit_obj$fl$elbo
```

## Visualizations related to fit

This is the true Gram matrix, $\frac{1}{p}XX'$:
```{r}
plot_heatmap(group_nonoverlap_1$data$YYt, colors = c('blue','gray96', 'red'), brks = seq(-max(abs(group_nonoverlap_1$data$YYt)), max(abs(group_nonoverlap_1$data$YYt)), length.out = 50))
```

This is a heatmap of the estimate of the Gram matrix, $\hat{L}\hat{L}'$:
```{r}
plot_heatmap(group_nonoverlap_1_ebmfcov_diag_1$est_LLt, brks = seq(0, max(group_nonoverlap_1_ebmfcov_diag_1$est_LLt), length.out = 50))
```

This is a scatter plot of the fitted values vs observed values:
```{r}
ggplot(data = NULL, aes(x = c(group_nonoverlap_1$data$YYt), y = c(group_nonoverlap_1_ebmfcov_diag_1$est_LLt))) + geom_point() + ylim(-1, 6) + xlim(-1,6) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

The model that we fit is $XX' \approx LL' + D$, so let's focus on the off-diagonal elements for now.
```{r}
diag_idx <- seq(1, prod(dim(group_nonoverlap_1$data$YYt)), length.out = ncol(group_nonoverlap_1$data$YYt))
off_diag_idx <- setdiff(c(1:prod(dim(group_nonoverlap_1$data$YYt))), diag_idx) 
```

```{r}
ggplot(data = NULL, aes(x = c(group_nonoverlap_1$data$YYt)[off_diag_idx], y = c(group_nonoverlap_1_ebmfcov_diag_1$est_LLt)[off_diag_idx])) + geom_point() + ylim(-1, 6) + xlim(-1,6) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

## Observations
Based off the visualizations, it seems like this method wants to add an intercept term. This is not totally surprising since the loadings and factors are added in a greedy way, and for this scenario, an intercept factor provides an equally good rank one fit as a factor identifying one group. However, looking at the heatmap of $LL'$, this looks like a bad estimate for the Gram matrix. (Side note: it's possible that if we add another factor, this fifth factor would pick up the last group.)

We wonder if backfitting would fix this issue because if you took out the effect of the three groups, then the intercept factor would be a bad fit to the residualized data matrix. We will test this in the next section.

# Adding backfitting
Now, we add a backfitting step whenever we fit the flash fits.
```{r}
cov_fit_backfit <- function(covmat, ebnm_fn = ebnm::ebnm_point_laplace, Kmax = 1000, verbose.lvl = 0, backfit_iter = 500, backfit_tol = NULL, cov_fit_init = NULL) {
  if (is.null(cov_fit_init)){
    fl <- flash_init(covmat, var_type = 0) %>%
      flash_set_verbose(verbose.lvl) %>%
      flash_greedy(ebnm_fn = ebnm_fn, Kmax = Kmax) %>%
      flash_backfit(maxiter = backfit_iter, tol = backfit_tol)
  }
  else{
    fl <- flash_init(covmat, var_type = 0) %>%
      flash_factors_init(cov_fit_init)
  }
  s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
  s2_diff <- Inf
  while(s2 > 0 && abs(s2_diff - 1) > 1e-4) {
    covmat_minuss2 <- covmat - diag(rep(s2, ncol(covmat)))
    fl <- flash_init(covmat_minuss2, var_type = 0) %>%
      flash_set_verbose(verbose.lvl) %>%
      flash_greedy(ebnm_fn = ebnm_fn, Kmax = Kmax) %>%
      flash_backfit(maxiter = backfit_iter, tol = backfit_tol)
    old_s2 <- s2
    s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
    s2_diff <- s2 / old_s2
  }
  
  return(list(fl=fl, s2 = s2))
}
```

```{r}
cov_fit_backfit_fit <- cov_fit_backfit(group_nonoverlap_1$data$YYt, ebnm_fn = ebnm::ebnm_generalized_binary, 
                                       Kmax = group_nonoverlap_1$data$K, backfit_iter = 500, backfit_tol = 10^(-16))
```

## Visualization of Estimate

This is a heatmap of $\hat{L}_{backfit}$, the estimate for $L$ from EBMFcov with backfitting (this is not scaled, so the scaling is arbitrary):
```{r}
plot_heatmap(cov_fit_backfit_fit$fl$L_pm, brks = seq(0, max(cov_fit_backfit_fit$fl$L_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}_{backfit}$ (this is not scaled, so the scaling is arbitrary):
```{r}
plot_loadings(cov_fit_backfit_fit$fl$L_pm, pops_vec)
```

This is a heatmap of $\hat{F}_{backfit}$, the estimate for $L$ from EBMFcov with backfitting (this is not scaled, so the scaling is arbitrary):
```{r}
plot_heatmap(cov_fit_backfit_fit$fl$F_pm, brks = seq(0, max(cov_fit_backfit_fit$fl$F_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{F}_{backfit}$ (this is not scaled, so the scaling is arbitrary):
```{r}
plot_loadings(cov_fit_backfit_fit$fl$F_pm, pops_vec)
```

This is the elbo of the flash fit:
```{r}
cov_fit_backfit_fit$fl$elbo
```

### Observations
We see that with the backfit, the estimate looks closer to what we want. This suggests that the issue with the original estimate is a computational issue (i.e. the objective function is not actively favoring the first estimate) that can be mitigated with backfitting. However, the recovered group effect is not as strong as the other group effects. I tried backfitting with larger numbers of iterations and backfitting with a decreased tolerance to see if the estimate improves (i.e. testing does the backfit have slow convergence?), but it doesn't seem to improve the estimate. After meeting with Matthew, he suggested I look at scaling. (The scaling issues also become more obvious when we also look at the estimate for $F$). So in the next section, I try re-scaling $L$ using the ldf function from flashier.

# Rescaling
It looks like there is an issue with scaling. So we try rescaling the estimates with the ldf function from flashier.
```{r}
scaled_cov_fit_backfit_fit <- ldf(cov_fit_backfit_fit$fl)
scaled_cov_fit_backfit_L <- scaled_cov_fit_backfit_fit$L %*% diag(sqrt(scaled_cov_fit_backfit_fit$D))
scaled_cov_fit_backfit_F <- scaled_cov_fit_backfit_fit$F %*% diag(sqrt(scaled_cov_fit_backfit_fit$D))
```

This is a heatmap of $\hat{L}_{backfit-scaled}$, the estimate for $L$ from EBMFcov with backfitting:
```{r}
plot_heatmap(scaled_cov_fit_backfit_L, brks = seq(0, max(scaled_cov_fit_backfit_L), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}_{backfit-scaled}$:
```{r}
plot_loadings(scaled_cov_fit_backfit_L, pops_vec)
```

This is a heatmap of $\hat{F}_{backfit-scaled}$, the estimate for $L$ from EBMFcov with backfitting:
```{r}
plot_heatmap(scaled_cov_fit_backfit_F, brks = seq(0, max(scaled_cov_fit_backfit_L), length.out = 50))
```

This is a scatterplot of the entries of $\hat{F}_{backfit-scaled}$:
```{r}
plot_loadings(scaled_cov_fit_backfit_F, pops_vec)
```

This is a heatmap of $\hat{L}_{backfit-scaled} \hat{L}_{backfit-scaled}'$:
```{r}
scaled_cov_fit_backfit_fit_estLLt <- tcrossprod(scaled_cov_fit_backfit_L)
```

```{r}
plot_heatmap(scaled_cov_fit_backfit_fit_estLLt, brks = seq(0, max(scaled_cov_fit_backfit_fit_estLLt), length.out = 50))
```

This is a scatter plot of the fitted values vs observed values:
```{r}
ggplot(data = NULL, aes(x = c(group_nonoverlap_1$data$YYt), y = c(scaled_cov_fit_backfit_fit_estLLt))) + geom_point() + ylim(-1, 7) + xlim(-1,7) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

This is a scatter plot of the off-diagonal entries of the fitted values vs those of the observed values:
```{r}
ggplot(data = NULL, aes(x = c(group_nonoverlap_1$data$YYt)[off_diag_idx], y = c(scaled_cov_fit_backfit_fit_estLLt)[off_diag_idx])) + geom_point() + ylim(-1, 7) + xlim(-1,7) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

## Observations
After rescaling, we see that the estimate for the covariance matrix is much better. All the group effects have the same scale, as expected. In addition, the fitted values now closely match the observed values. However, one strange feature of this estimate is there are a handful of zero loadings which become non-zero during the backfitting. My hypothesis is that as the baseline values in the first value go to zero, some of the zero values in the other loadings increase. Another thing I want to try is initializing the fit at the true values; I want to see if that estimate also has these small non-zero loadings.

# Progression of the backfitting

## Backfit after 3 iterations
```{r}
cov_fit_backfit_iter3_fit <- cov_fit_backfit(group_nonoverlap_1$data$YYt, ebnm_fn = ebnm::ebnm_generalized_binary, 
                                       Kmax = group_nonoverlap_1$data$K, backfit_iter = 3)
```

This is a scatterplot of the entries of $\hat{L}_{backfit}$ (this is not scaled, so the scaling is arbitrary):
```{r}
plot_loadings(cov_fit_backfit_iter3_fit$fl$L_pm, pops_vec)
```

This is a scatterplot of the entries of $\hat{F}_{backfit}$ (this is not scaled, so the scaling is arbitrary):
```{r}
plot_loadings(cov_fit_backfit_iter3_fit$fl$F_pm, pops_vec)
```

This is the elbo of the flash fit:
```{r}
cov_fit_backfit_iter3_fit$fl$elbo
```

## Backfit after 4 iterations
```{r}
cov_fit_backfit_iter4_fit <- cov_fit_backfit(group_nonoverlap_1$data$YYt, ebnm_fn = ebnm::ebnm_generalized_binary, 
                                       Kmax = group_nonoverlap_1$data$K, backfit_iter = 4)
```

This is a scatterplot of the entries of $\hat{L}_{backfit}$ (this is not scaled, so the scaling is arbitrary):
```{r}
plot_loadings(cov_fit_backfit_iter4_fit$fl$L_pm, pops_vec)
```

This is a scatterplot of the entries of $\hat{F}_{backfit}$ (this is not scaled, so the scaling is arbitrary):
```{r}
plot_loadings(cov_fit_backfit_iter4_fit$fl$F_pm, pops_vec)
```

This is the elbo of the flash fit:
```{r}
cov_fit_backfit_iter4_fit$fl$elbo
```

# Initialize at true values

We run EBMFcov initialized with the true loadings matrix:
```{r}
cov_fit_true_init_fit <- cov_fit_backfit(group_nonoverlap_1$data$YYt, ebnm_fn = ebnm::ebnm_generalized_binary, 
                                       Kmax = group_nonoverlap_1$data$K, backfit_iter = 500, backfit_tol = 10^(-16), cov_fit_init = list(group_nonoverlap_1$data$LL, group_nonoverlap_1$data$LL))
```

We rescale the estimates such that $XX' = LL'$ (the scaling from EBMFcov is arbitrary):
```{r}
scaled_cov_fit_true_init_fit <- ldf(cov_fit_true_init_fit$fl)
scaled_cov_fit_true_init_L <- scaled_cov_fit_true_init_fit$L %*% diag(sqrt(scaled_cov_fit_true_init_fit$D))
scaled_cov_fit_true_init_F <- scaled_cov_fit_true_init_fit$F %*% diag(sqrt(scaled_cov_fit_true_init_fit$D))
```

## Visualization of Estimate

This is a heatmap of $\hat{L}_{true-init}$, the estimate for $L$ from EBMFcov intialized with the true values:
```{r}
plot_heatmap(scaled_cov_fit_true_init_L, brks = seq(0, max(scaled_cov_fit_true_init_L), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}_{true-init}$ (this is not scaled, so the scaling is arbitrary):
```{r}
plot_loadings(scaled_cov_fit_true_init_L, pops_vec)
```

This is a heatmap of $\hat{F}_{backfit}$, the estimate for $L$ from EBMFcov with backfitting (this is not scaled, so the scaling is arbitrary):
```{r}
plot_heatmap(scaled_cov_fit_true_init_F, brks = seq(0, max(scaled_cov_fit_true_init_F), length.out = 50))
```

This is a scatterplot of the entries of $\hat{F}_{backfit}$ (this is not scaled, so the scaling is arbitrary):
```{r}
plot_loadings(scaled_cov_fit_true_init_F, pops_vec)
```

This is the elbo of the flash fit:
```{r}
cov_fit_true_init_fit$fl$elbo
```

## Observations
The estimate from EBMFcov initialized with the true values also has these small non-zero values. This suggests that the objective function favors an estimate with these small loading values rather than zero values. I'm not sure if this is related to the data?

# Try more factors
Another question of interest is whether adding a fifth factor would add the final group effect. Furthermore, if we do backfit, then it would make sense for the first factor to zero out (since it is not necessary to explain the fit anymore). We test this out in this section.

We run EBMFcov with `Kmax = 5`:
```{r}
cov_fit_backfit_K5_fit <- cov_fit_backfit(group_nonoverlap_1$data$YYt, ebnm_fn = ebnm::ebnm_generalized_binary, 
                                       Kmax = 5, backfit_iter = 500)
```

This is a scatterplot of the entries of $\hat{L}_{backfit-K5}$ (this is not scaled, so the scaling is arbitrary):
```{r}
plot_loadings(cov_fit_backfit_K5_fit$fl$L_pm, pops_vec)
```

This is a scatterplot of the entries of $\hat{F}_{backfit-K5}$ (this is not scaled, so the scaling is arbitrary):
```{r}
plot_loadings(cov_fit_backfit_K5_fit$fl$F_pm, pops_vec)
```

This is the elbo of the flash fit:
```{r}
cov_fit_backfit_K5_fit$fl$elbo
```

## Observations
The estimate from fitting EBMFcov with a higher Kmax still has four factors. During the greedy algorithm, when the fifth factor is added, flash states the "Factor doesn't significantly increase objective and won't be added". I think the structure of the Gram matrix strongly supports the matrix having rank four, so perhaps flash doesn't want to add another factor, creating an over-complete basis.

