---
title: "tree_ebmfcov_diag"
author: "Annie Xie"
date: "2025-02-19"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Introduction
In this analysis, we explore EBMFcov in the balanced tree setting.

## Questions of Interest
This analysis is complementary to other analyses of methods in the tree setting. One reason I want to explore EBMFcov in this setting is because GBCD does very well in this setting. GBCD and EBMFcov are very similar methods -- they both apply flashier in the covariance space. However, EBMFcov does not perform as well in the tree setting. By exploring EBMFcov, I'm hoping to gain insights into both methods.

```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(pheatmap)
library(flashier)
```

```{r}
source('code/visualization_functions.R')
```

```{r}
compute_L2_fit <- function(est, dat, with_diag = FALSE){
  if (with_diag == FALSE){
    score <- sum((dat - est)^2) - sum((diag(dat) - diag(est))^2)
  }
  else{
    score <- sum((dat - est)^2)
  }
  return(score)
}
```

# EBMFcov (without backfitting)

```{r}
baltree_4pop_1 <- readRDS("data/baltree_4pop_1.rds")
baltree_4pop_1_ebmfcov_1 <- readRDS("data/baltree_4pop_1_ebmfcov_diag_1.rds")
```

## Visualization of Estimate
This is a heatmap of the true $L$ that we are hoping to recover:
```{r}
plot_heatmap(baltree_4pop_1$true_L)
```

This is a heatmap of $\hat{L}$, the estimate for $L$ from EBMFcov (this estimate is not scaled):
```{r}
plot_heatmap(baltree_4pop_1_ebmfcov_1$est_L, brks = seq(0, max(baltree_4pop_1_ebmfcov_1$est_L), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(baltree_4pop_1_ebmfcov_1$est_L, pops_vec)
```

This is a heatmap of $\hat{F}$ (we hope this is similar to $\hat{L}$, up to a scaling):
```{r}
plot_heatmap(baltree_4pop_1_ebmfcov_1$fit_obj$fl$F_pm, brks = seq(0, max(baltree_4pop_1_ebmfcov_1$fit_obj$fl$F_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{F}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(baltree_4pop_1_ebmfcov_1$fit_obj$fl$F_pm, pops_vec)
```

This is the estimate of the diagonal component:
```{r}
baltree_4pop_1_ebmfcov_1$fit_obj$s2
```

This is the estimate of the residual noise variance:
```{r}
baltree_4pop_1_ebmfcov_1$fit_obj$fl$residuals_sd
```

## Visualizations related to fit

This is the true Gram matrix, $\frac{1}{p}XX'$:
```{r}
plot_heatmap(baltree_4pop_1$data$YYt, colors = c('blue','gray96', 'red'), brks = seq(-max(abs(baltree_4pop_1$data$YYt)), max(abs(baltree_4pop_1$data$YYt)), length.out = 50))
```

This is a heatmap of the estimate of the Gram matrix, $\hat{L}\hat{L}'$ (the $\hat{L}$ used here is scaled):
```{r}
plot_heatmap(baltree_4pop_1_ebmfcov_1$est_LLt, brks = seq(0, max(baltree_4pop_1_ebmfcov_1$est_LLt), length.out = 50))
```

This is a scatter plot of the fitted values vs observed values:
```{r}
ggplot(data = NULL, aes(x = c(baltree_4pop_1$data$YYt), y = c(baltree_4pop_1_ebmfcov_1$est_LLt))) + geom_point() + ylim(-1, 18) + xlim(-1,18) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

Let's focus on the off-diagonal elements:
```{r}
diag_idx <- seq(1, prod(dim(baltree_4pop_1$data$YYt)), length.out = ncol(baltree_4pop_1$data$YYt))
off_diag_idx <- setdiff(c(1:prod(dim(baltree_4pop_1$data$YYt))), diag_idx) 
```

```{r}
ggplot(data = NULL, aes(x = c(baltree_4pop_1$data$YYt)[off_diag_idx], y = c(baltree_4pop_1_ebmfcov_1$est_LLt)[off_diag_idx])) + geom_point() + ylim(-1, 18) + xlim(-1,18) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

This is the L2-norm of the difference (not including the diagonal entries):
```{r}
compute_L2_fit(baltree_4pop_1_ebmfcov_1$est_LLt, baltree_4pop_1$data$YYt)
```

This is the elbo of the final flash fit:
```{r}
baltree_4pop_1_ebmfcov_1$fit_obj$fl$elbo
```

## Observations
EBMFcov only returns three factors -- one intercept factor and two factors corresponding to the two different subtypes. (I checked the EBMFcov estimates for all ten simulated tree datasets, and all the estimates had this structure). As a result, the estimate for the Gram matrix $\hat{L} \hat{L}'$, is missing the four block components corresponding to group-specific effects. I'm not exactly sure where it stops adding new factors after three factors. Maybe the group-specific effects are getting lumped in with the residual noise?

# Try initializing with true values
In this section, I try initializing EBMFcov with the true values and see if it returns an estimate similar to the true values.

```{r}
cov_fit_true_init <- function(covmat, true_L, ebnm_fn = ebnm::ebnm_point_laplace, Kmax = 1000, verbose.lvl = 0) {
  fl <- flash_init(covmat, var_type = 0) %>%
    flash_factors_init(list(true_L, true_L), ebnm_fn = ebnm_fn)
  s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
  s2_diff <- Inf
  while(s2 > 0 && abs(s2_diff - 1) > 1e-4) {
    covmat_minuss2 <- covmat - diag(rep(s2, ncol(covmat)))
    fl <- flash_init(covmat_minuss2, var_type = 0) %>%
      flash_set_verbose(verbose.lvl) %>%
      flash_greedy(ebnm_fn = ebnm_fn, Kmax = Kmax)
    old_s2 <- s2
    s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
    s2_diff <- s2 / old_s2
  }
  return(list(fl = fl, s2 = s2))
}
```

```{r}
ebmfcov_true_init <- cov_fit_true_init(baltree_4pop_1$data$YYt, baltree_4pop_1$data$LL, ebnm_fn = ebnm::ebnm_generalized_binary, verbose.lvl = 2)
```

## Visualization of Estimate
This is a heatmap of $\hat{L}_{true-init}$:
```{r}
plot_heatmap(ebmfcov_true_init$fl$L_pm, brks = seq(0, max(ebmfcov_true_init$fl$L_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}_{true-init}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(ebmfcov_true_init$fl$L_pm, pops_vec)
```

This is a heatmap of $\hat{F}_{true-init}$:
```{r}
plot_heatmap(ebmfcov_true_init$fl$F_pm, brks = seq(0, max(ebmfcov_true_init$fl$F_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{F}_{true-init}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(ebmfcov_true_init$fl$F_pm, pops_vec)
```

## Visualizations related to fit

```{r}
scaled_ebmfcov_true_init <- ldf(ebmfcov_true_init$fl)
scaled_ebmfcov_true_init_L <- scaled_ebmfcov_true_init$L %*% diag(sqrt(scaled_ebmfcov_true_init$D))
scaled_ebmfcov_true_init_estLLt <- tcrossprod(scaled_ebmfcov_true_init_L)
```

This is a heatmap of the estimate of the Gram matrix, $\hat{L}_{true-init}\hat{L}_{true-init}'$ (the $\hat{L}_{true-init}$ used here is scaled):
```{r}
plot_heatmap(scaled_ebmfcov_true_init_estLLt, brks = seq(0, max(scaled_ebmfcov_true_init_estLLt), length.out = 50))
```

This is a scatter plot of the fitted values vs observed values:
```{r}
ggplot(data = NULL, aes(x = c(baltree_4pop_1$data$YYt), y = c(scaled_ebmfcov_true_init_estLLt))) + geom_point() + ylim(-1, 18) + xlim(-1,18) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

Let's focus on the off-diagonal elements:
```{r}
diag_idx <- seq(1, prod(dim(baltree_4pop_1$data$YYt)), length.out = ncol(baltree_4pop_1$data$YYt))
off_diag_idx <- setdiff(c(1:prod(dim(baltree_4pop_1$data$YYt))), diag_idx) 
```

```{r}
ggplot(data = NULL, aes(x = c(baltree_4pop_1$data$YYt)[off_diag_idx], y = c(scaled_ebmfcov_true_init_estLLt)[off_diag_idx])) + geom_point() + ylim(-1, 18) + xlim(-1,18) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

This is the L2-norm of the difference (not including the diagonal entries):
```{r}
compute_L2_fit(scaled_ebmfcov_true_init_estLLt, baltree_4pop_1$data$YYt)
```

This is the elbo of the final flash fit:
```{r}
ebmfcov_true_init$fl$elbo
```

## Observations
Even when initializing with the true values, EBMFcov yielded the same estimate. This was pretty surprising to me because GBCD initialized with the true values returns a loadings estimate that looks like a tree. Given the similarity between the methods, I expected EBMFcov to do the same. Therefore, I more closely compared the GBCD code with the EBMFcov code.

# Comparing GBCD to EBMFcov
Both EBMFcov and GBCD alternate between estimating the diagonal component $s^2$ and fitting flash to $XX' - s^2I_n$. In the iterative process, when EBMFcov is fitting flash to $XX' - s^2I_n$, it creates a new flash object from scratch; it initializes the flash object and does the greedy procedure (and can potentially backfit). On the other hand, GBCD maintains the flash object and instead updates the data in each iteration. Therefore, I edited the EBMFcov to match the GBCD updates. In this code, we also initialize from the true values.

```{r}
cov_fit_true_init_v2 <- function(covmat, true_L, ebnm_fn = ebnm::ebnm_point_laplace, Kmax = 1000, verbose.lvl = 0) {
  fl <- flash_init(covmat, var_type = 0) %>%
    flash_factors_init(list(true_L, true_L), ebnm_fn = ebnm_fn)
  s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
  s2_diff <- Inf
  while(s2 > 0 && abs(s2_diff - 1) > 1e-4) {
    covmat_minuss2 <- covmat - diag(rep(s2, ncol(covmat)))
    fl <- fl %>%
      flash_update_data(covmat_minuss2) %>%
      flash_set_verbose(verbose.lvl) %>%
      flash_backfit()
    old_s2 <- s2
    s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
    s2_diff <- s2 / old_s2
  }
  
  return(list(fl=fl, s2 = s2))
}
```

```{r}
ebmfcov_true_init2 <- cov_fit_true_init_v2(baltree_4pop_1$data$YYt, baltree_4pop_1$data$LL, ebnm_fn = ebnm::ebnm_generalized_binary, verbose.lvl = 2)
```

## Visualization of Estimate
This is a heatmap of $\hat{L}_{true-init2}$:
```{r}
plot_heatmap(ebmfcov_true_init2$fl$L_pm, brks = seq(0, max(ebmfcov_true_init2$fl$L_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}_{true-init2}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(ebmfcov_true_init2$fl$L_pm, pops_vec)
```

This is a heatmap of $\hat{F}_{true-init2}$:
```{r}
plot_heatmap(ebmfcov_true_init2$fl$F_pm, brks = seq(0, max(ebmfcov_true_init2$fl$F_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{F}_{true-init2}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(ebmfcov_true_init2$fl$F_pm, pops_vec)
```

## Visualizations related to fit

```{r}
scaled_ebmfcov_true_init2 <- ldf(ebmfcov_true_init2$fl)
scaled_ebmfcov_true_init2_L <- scaled_ebmfcov_true_init2$L %*% diag(sqrt(scaled_ebmfcov_true_init2$D))
scaled_ebmfcov_true_init2_estLLt <- tcrossprod(scaled_ebmfcov_true_init2_L)
```

This is a heatmap of the estimate of the Gram matrix, $\hat{L}_{true-init2}\hat{L}_{true-init2}'$ (the $\hat{L}_{true-init2}$ used here is scaled):
```{r}
plot_heatmap(scaled_ebmfcov_true_init2_estLLt, brks = seq(0, max(scaled_ebmfcov_true_init2_estLLt), length.out = 50))
```

This is a scatter plot of the fitted values vs observed values:
```{r}
ggplot(data = NULL, aes(x = c(baltree_4pop_1$data$YYt), y = c(scaled_ebmfcov_true_init2_estLLt))) + geom_point() + ylim(-1, 18) + xlim(-1,18) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

Let's focus on the off-diagonal elements:
```{r}
diag_idx <- seq(1, prod(dim(baltree_4pop_1$data$YYt)), length.out = ncol(baltree_4pop_1$data$YYt))
off_diag_idx <- setdiff(c(1:prod(dim(baltree_4pop_1$data$YYt))), diag_idx) 
```

```{r}
ggplot(data = NULL, aes(x = c(baltree_4pop_1$data$YYt)[off_diag_idx], y = c(scaled_ebmfcov_true_init2_estLLt)[off_diag_idx])) + geom_point() + ylim(-1, 18) + xlim(-1,18) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

This is the L2-norm of the difference (not including the diagonal entries):
```{r}
compute_L2_fit(scaled_ebmfcov_true_init2_estLLt, baltree_4pop_1$data$YYt)
```

This is the elbo of the final flash fit:
```{r}
ebmfcov_true_init2$fl$elbo
```

## Observations
This new version of EBMFcov initialized with the true values does yield an estimate that looks like a tree. There's one intercept-like factor, two subtype factors, and four group effect factors. The estimate $\hat{L}_{true-init2} \hat{L}_{true-init2}'$ is much closer to the observed Gram matrix. Furthermore, the elbo for the final flash fit is higher. This suggests that the objective function prefers this solution to the one originally found by EBMFcov. In addition, the change from using the greedy procedure to just updating the data matrix makes me think that the greedy procedure generally does not perform well in this setting.

# Testing EBMFcov on a larger tree
Previously, I mentioned that maybe flashier was lumping the group effects into the residual noise. Therefore, I wanted to test EBMFcov on a larger tree to see if it picks up the other subtype effects. In this section, we simulate a balanced 8 population tree

```{r}
cov_fit <- function(covmat, ebnm_fn = ebnm::ebnm_point_laplace, Kmax = 1000, verbose.lvl = 0) {
  fl <- flash_init(covmat, var_type = 0) %>%
    flash_set_verbose(verbose.lvl) %>%
    flash_greedy(ebnm_fn = ebnm_fn, Kmax = Kmax)
  s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
  s2_diff <- Inf
  while(s2 > 0 && abs(s2_diff - 1) > 1e-4) {
    covmat_minuss2 <- covmat - diag(rep(s2, ncol(covmat)))
    fl <- flash_init(covmat_minuss2, var_type = 0) %>%
      flash_set_verbose(verbose.lvl) %>%
      flash_greedy(ebnm_fn = ebnm_fn, Kmax = Kmax)
    old_s2 <- s2
    s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
    s2_diff <- s2 / old_s2
  }
  
  return(list(fl=fl, s2 = s2))
}
```

```{r}
sim_8pops <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  
  FF <- matrix(rnorm(15 * p, sd = rep(args$branch_sds, each = p)), ncol = 15)
  # if (args$constrain_F) {
  #   FF_svd <- svd(FF)
  #   FF <- FF_svd$u
  #   FF <- t(t(FF) * branch_sds * sqrt(p))
  # }
  
  LL <- matrix(0, nrow = n, ncol = 15)
  LL[, 1] <- 1
  LL[, 2] <- rep(c(1, 1, 1, 1, 0, 0, 0, 0), times = args$pop_sizes)
  LL[, 3] <- rep(c(0, 0, 0, 0, 1, 1, 1, 1), times = args$pop_sizes)
  LL[, 4] <- rep(c(1, 1, 0, 0, 0, 0, 0, 0), times = args$pop_sizes)
  LL[, 5] <- rep(c(0, 0, 1, 1, 0, 0, 0, 0), times = args$pop_sizes)
  LL[, 6] <- rep(c(0, 0, 0, 0, 1, 1, 0, 0), times = args$pop_sizes)
  LL[, 7] <- rep(c(0, 0, 0, 0, 0, 0, 1, 1), times = args$pop_sizes)
  LL[, 8] <- rep(c(1, 0, 0, 0, 0, 0, 0, 0), times = args$pop_sizes)
  LL[, 9] <- rep(c(0, 1, 0, 0, 0, 0, 0, 0), times = args$pop_sizes)
  LL[, 10] <- rep(c(0, 0, 1, 0, 0, 0, 0, 0), times = args$pop_sizes)
  LL[, 11] <- rep(c(0, 0, 0, 1, 0, 0, 0, 0), times = args$pop_sizes)
  LL[, 12] <- rep(c(0, 0, 0, 0, 1, 0, 0, 0), times = args$pop_sizes)
  LL[, 13] <- rep(c(0, 0, 0, 0, 0, 1, 0, 0), times = args$pop_sizes)
  LL[, 14] <- rep(c(0, 0, 0, 0, 0, 0, 1, 0), times = args$pop_sizes)
  LL[, 15] <- rep(c(0, 0, 0, 0, 0, 0, 0, 1), times = args$pop_sizes)
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
sim_args <- list(seed = 1, pop_sizes = rep(40,8), n_genes = 1000, branch_sds = rep(2,15), indiv_sd = 1)
baltree_8pop_1 <- sim_8pops(sim_args)
```

```{r}
baltree_8pop_1_ebmfcov_1 <- cov_fit(baltree_8pop_1$YYt, ebnm::ebnm_generalized_binary, Kmax = 15, verbose.lvl = 2)
```

## Visualization of Estimate
This is a heatmap of the true $L$ that we are hoping to recover:
```{r}
plot_heatmap(baltree_8pop_1$LL)
```

This is a heatmap of $\hat{L}_{8pop}$:
```{r}
plot_heatmap(baltree_8pop_1_ebmfcov_1$fl$L_pm, brks = seq(0, max(baltree_8pop_1_ebmfcov_1$fl$L_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{L}_{8pop}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40), rep('E', 40), rep('F', 40), rep('G', 40), rep('H', 40))
plot_loadings(baltree_8pop_1_ebmfcov_1$fl$L_pm, pops_vec)
```

This is a heatmap of $\hat{F}_{8pop}$:
```{r}
plot_heatmap(baltree_8pop_1_ebmfcov_1$fl$F_pm, brks = seq(0, max(baltree_8pop_1_ebmfcov_1$fl$F_pm), length.out = 50))
```

This is a scatterplot of the entries of $\hat{F}_{8pop}$:
```{r}
pops_vec <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40), rep('E', 40), rep('F', 40), rep('G', 40), rep('H', 40))
plot_loadings(baltree_8pop_1_ebmfcov_1$fl$F_pm, pops_vec)
```

## Visualizations related to fit

This is the true Gram matrix, $\frac{1}{p}XX'$:
```{r}
plot_heatmap(baltree_8pop_1$YYt, colors = c('blue','gray96', 'red'), brks = seq(-max(abs(baltree_8pop_1$YYt)), max(abs(baltree_8pop_1$YYt)), length.out = 50))
```

```{r}
scaled_baltree_8pop_ebmfcov_fit <- ldf(baltree_8pop_1_ebmfcov_1$fl)
scaled_baltree_8pop_ebmfcov_fit_L <- scaled_baltree_8pop_ebmfcov_fit$L %*% diag(sqrt(scaled_baltree_8pop_ebmfcov_fit$D))
scaled_baltree_8pop_ebmfcov_fit_estLLt <- tcrossprod(scaled_baltree_8pop_ebmfcov_fit_L)
```

This is a heatmap of the estimate of the Gram matrix, $\hat{L}_{8pop}\hat{L}_{8pop}'$ (the $\hat{L}_{8pop}$ used here is scaled):
```{r}
plot_heatmap(scaled_baltree_8pop_ebmfcov_fit_estLLt, brks = seq(0, max(scaled_baltree_8pop_ebmfcov_fit_estLLt), length.out = 50))
```

This is a scatter plot of the fitted values vs observed values:
```{r}
ggplot(data = NULL, aes(x = c(baltree_8pop_1$YYt), y = c(scaled_baltree_8pop_ebmfcov_fit_estLLt))) + geom_point() + ylim(-1, 20) + xlim(-1,20) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

Let's focus on the off-diagonal elements:
```{r}
diag_idx <- seq(1, prod(dim(baltree_8pop_1$YYt)), length.out = ncol(baltree_8pop_1$YYt))
off_diag_idx <- setdiff(c(1:prod(dim(baltree_8pop_1$YYt))), diag_idx) 
```

```{r}
ggplot(data = NULL, aes(x = c(baltree_8pop_1$YYt)[off_diag_idx], y = c(scaled_baltree_8pop_ebmfcov_fit_estLLt)[off_diag_idx])) + geom_point() + ylim(-1, 18) + xlim(-1,18) + xlab('Observed Values') + ylab('Fitted Values') + geom_abline(slope = 1, intercept = 0, color = 'red')
```

This is the L2-norm of the difference (not including the diagonal entries):
```{r}
compute_L2_fit(scaled_baltree_8pop_ebmfcov_fit_estLLt, baltree_8pop_1$YYt)
```

This is the elbo of the final flash fit:
```{r}
baltree_8pop_1_ebmfcov_1$fl$elbo
```

## Observations
In this setting, EBMFcov also returns an estimate with only three factors -- one intercept factor and two factors corresponding to the largest subtypes. In another analysis file, I explore the greedy procedure on tree data. Hopefully that exploration can give us insight into why EBMFcov only fits three factors in this setting.

